Ownership types in theory and practice (in Rust)

### 2022-11-30 paper.tex
Finished ownership transfer for heap based values

\usepackage{booktabs} % For formal tables
\usepackage{subcaption}
\usepackage{tikz}
\usepackage{pgfplots}
\usepackage{pgfplotstable}
\usepackage{hyphenat}
\usepackage{todonotes}
\usepackage[babel]{csquotes}
\usepackage{listings}
\lstset{frame=tb,
  language=Rust,
  aboveskip=3mm,
  belowskip=3mm,
  showstringspaces=false,
  columns=flexible,
  basicstyle={\small\ttfamily},
  numbers=none,
  numberstyle=\tiny\color{gray},
  keywordstyle=\color{blue},
  commentstyle=\color{dkgreen},
  stringstyle=\color{mauve},
  breaklines=true,
  breakatwhitespace=true,
  tabsize=3
}

@@

The following describes the concepts of how variables and memory interact in Rust.
\cite{rust-book}

\subsubsection{Move}

% TODO: wording of value and types used interchangedly here, maybe incorrect
% First, one must differentiate between values that are stored on the stack and those stored on the heap.
In Rust, primitive data types (e.g. signed 32 bit integers \emph{i32}) are immutable by default and stored on the stack, since their sizes are small and known at compile time.
Values are stored on the heap if their size is dynamic and therefore not known at compile time.

% TODO: a good example of what?
The different representations of strings in Rust provide a good example.
Rust supports immutable string literals \emph{\&str}, which are stored on the stack, and objects of type \emph{String}, which are stored as a vector of bytes on the heap and are, therefore, growable. \cite{rust-by-example}

% TODO: footnote that main class etc. have been omitted for sake of simplicity
The following piece of code demonstrates how primitive values are \emph{moved} into other variables in Rust.
\begin{lstlisting}
let x: &str = "hello world";
let y: &str = x;
println!("x: {}, y: {}", x, y);
\end{lstlisting}
Since \emph{x} has a known, fixed size, its value is simply copied into \emph{y}.
Therefore, the program compiles and, as expected, "x: hello world, y: hello world" is printed to stdout.
% TODO: explain what Rust traits are
Internally, these primitive types are annotated with the \emph{Copy} trait, which specifies that

However, a similar code snippet that replaces the string literal with the \emph{String} type behaves differently.
\begin{lstlisting}
let x: String = String::from("hello world");
let y: String = x;
println!("x: {}, y: {}", x, y);
\end{lstlisting}
% TODO: go into detail on how strings are stored on heap.
The \emph{String} value stored in \emph{x} could be arbitrarily long at runtime.
Thus, copying the whole string into \emph{y} could be very inefficient.
% TODO: avoid and at beginning of sentence
And, since the \emph{String} value can only have one owner, it is \emph{moved} into \emph{y}, which will become its new owner.
After the move, \emph{x} is no longer valid.
Because Rust prevents you from using the invalid reference to \emph{x} in the third line, the program will not compile.

% TODO: add section explaing borrowing, since it's related ot move

\subsubsection{Clone}

@@

>>>>

\usepackage{booktabs} % For formal tables
\usepackage{subcaption}
\usepackage{tikz}
\usepackage{pgfplots}
\usepackage{pgfplotstable}
\usepackage{hyphenat}
\usepackage{todonotes}
\usepackage[babel]{csquotes}
\usepackage{listings}
\lstset{frame=tb,
  language=Rust,
  aboveskip=3mm,
  belowskip=3mm,
  showstringspaces=false,
  columns=flexible,
  basicstyle={\small\ttfamily},
  numbers=left,
  numberstyle=\tiny\color{gray},
  keywordstyle=\color{blue},
  commentstyle=\color{dkgreen},
  stringstyle=\color{mauve},
  breaklines=true,
  breakatwhitespace=true,
  tabsize=3
}

@@

% TODO: source
In addition to providing the memory safety that ownership types provide, ownership transfers can also reduce redundant copy operations.
Since heap-based values could be arbitrarily large at runtime, copying the whole value instead of transferring ownership could be very inefficient.

The following describes how the concept of ownership is implemented in Rust.

% TODO: wording of value and types used interchangedly here, maybe incorrect
% TODO: maybe add defintion of ownership transfer if not obvious
\subsubsection{Transferring ownership}

First, one must differentiate between values that are stored on the stack and those stored on the heap.
In Rust, primitive data types (e.g. signed 32 bit integers \emph{i32}) are immutable by default and stored on the stack, since their sizes are small and known at compile time.
In contrast, values are stored on the heap if their size is dynamic and, therefore, unknown at compile time.
The ownership concepts in Rust mostly apply to values stored on the heap.
% Thus, the behaviour of values stored on the stack will be explained later.

\paragraph{Heap-based variables}

There are several different scenarios in which ownership is transferred.

In the simplest scenario, ownership is transferred from the variable \emph{a} to the variable \emph{b} when \emph{b} is assigned to \emph{a}.
\begin{lstlisting}
let a = HeapBasedType::default();
let b = a;
\end{lstlisting}
In Rust terminology, \emph{a} is \emph{moved} into \emph{b}.
After the move, \emph{a} is no longer valid and \emph{b} is the new owner of the value.

Heap-based values will always be generated by a function at runtime.
Therefore, one can generalize the above scenario to ownership being transferred from the variable \emph{a} to the variable \emph{b} when \emph{b} is assigned to the return value \emph{a} of \emph{foo}.
\begin{lstlisting}
fn main() {
  let b = foo();
}
fn foo() {
  let a = HeapBasedType::default();
  a
}
\end{lstlisting}

Lastly, ownership is transferred from the variable \emph{a} to the variable \emph{b} when \emph{a} is passed to function \emph{foo} which takes a parameter \emph{b}.
\begin{lstlisting}
fn main() {
  let a = HeapBasedType::default();
  foo(a);
}
fn foo(b: HeapBasedType) {
  // do something with b
}
\end{lstlisting}
In \emph{foo}, \emph{b} is the owner of the value until it is \emph{dropped} at the end of the function.
Accessing the previous owner of a value after the ownership of that value has been transferred elsewhere will lead to a compile-time error.
In this case, trying to access \emph{a} after line 3 would not compile, as \emph{a} is no longer the owner of the value.

% TODO: avoid and at beginning of sentence
% And, since the \emph{String} value can only have one owner, it is \emph{moved} into \emph{y}, which will become its new owner.
% Because Rust prevents you from using the invalid reference to \emph{x} in the third line, the program will not compile.

% The two main representations of strings in Rust provide a good example.
% Rust supports immutable string literals \emph{\&str}, which are stored on the stack, and objects of type \emph{String}, which are stored as a vector of bytes on the heap and are, therefore, growable. \cite{rust-by-example}

% TODO: bad sentence
% The following piece of code shows a \emph{String} value being \emph{moved} from one variable into another.
The heap-allocated \emph{String} type in Rust provides a good example to demonstrate the transfer of ownership of values.
\begin{lstlisting}
let x: String = String::from("hello world");
let y: String = x;
println!("x: {}, y: {}", x, y);
\end{lstlisting}
% TODO: go into detail on how strings/heap types are stored on heap.
The \emph{String} value stored in \emph{x} could be arbitrarily long at runtime.
Thus, copying the whole string into \emph{y} could be very inefficient.
% TODO: avoid and at beginning of sentence
And, since the \emph{String} value can only have one owner, it is \emph{moved} into \emph{y}, which will become its new owner.
After the move, \emph{x} is no longer valid.
Because Rust prevents you from using the invalid reference to \emph{x} in the third line, the program will not compile.

@@

The ownership of a value is not only transferred during the assignment to another variable, but also when it is passed to a function as a parameter and when it is returned from a function.

\subsubsection{References/Aliases}

\subsubsection{Borrowing}

\subsubsection{Slice type}

\subsubsection{Stack-based values}

First, one must differentiate between values that are stored on the stack and those stored on the heap.
% TODO: leave out immutable, adds no meaning to sentence
In Rust, primitive data types (e.g. signed 32 bit integers \emph{i32}) are immutable by default and stored on the stack, since their sizes are small and known at compile time.

% TODO: a good example of what?

% TODO: footnote that main class etc. have been omitted for sake of simplicity
% However, a similar code snippet that replaces the string literal with the \emph{String} type behaves differently.
% The following piece of code demonstrates how primitive values are \emph{moved} into other variables in Rust.
\begin{lstlisting}
let x: &str = "hello world";
let y: &str = x;
println!("x: {}, y: {}", x, y);
\end{lstlisting}
Since \emph{x} has a known, fixed size, its value is simply copied into \emph{y}.
Therefore, the program compiles and, as expected, "x: hello world, y: hello world" is printed to stdout.
% TODO: explain what Rust traits are
Internally, these primitive types are annotated with the \emph{Copy} trait, which specifies that variables are trivially copied instead of moved.
